Creating More Meaningful Explanations for Audio Deepfake Detection

### 2025-03-05 draft.tex
update citations to google scholar format
transfer edits from the conference version of the paper

\documentclass{article}
\usepackage{graphicx} % Required for inserting images
\usepackage{biblatex} % Imports biblatex package
\addbibresource{references.bib}
\usepackage{changepage}
\usepackage{parskip}

@@

    {\begin{table}[h]
			\begin{tabular}{lll}
				{Erstprüfer:} & \erstpruefer & \erstprueferMail \\
			\end{tabular}
		\end{table}}     
		\vfill
	\end{titlepage}
    \newpage
    \section{Abstract}
	\sloppy
	With the unprecedented advancement of Generative Artificial Intelligence (GenAI), the threat
	of voice scams using synthetic voices has become a serious concern across various sectors.
	Recent efforts have focused on identifying fake voices through handcrafted features, deep
	learning models, and hybrid approaches. However, most existing methods lack explainability,
	rendering their predictions non-transparent to users. This paper proposes a novel,
	interpretable, and transparent method for fake voice identification by introducing a hybrid
	deep learning model that leverages multiple extracted features. The hybrid model consists of
	two main components: the first component addresses heterogeneous feature spaces by employing
	deep convolutional sub-models tailored to individual features, while the second component, the
	terminus model, utilizes the concatenated representations from the final layers of each
	sub-model as input. The terminus model follows a typical multi-layer perceptron architecture,
	enabling effective integration and classification of the diverse feature representations. To
	enhance interpretability, we decompose the model’s decisions using Local Interpretable
	Model-agnostic Explanations (LIME), exploiting the identical feature representation before the
	concatenation layers to address challenges related to multi-dimensional feature
	representations. To evaluate the importance and trustworthiness of features in the generated
	explanations, we propose two metrics: importance and trust. Extensive experiments are
	conducted on the In-the-Wild dataset, which is designed to test the generalization capability
	of synthetic audio detection methods. The experimental results demonstrate that our approach
	achieves performance comparable to benchmark methods. Furthermore, the results based on our
	proposed metrics conclude that certain perceptible features demonstrate promise for generating
	explanations that are meaningful to general users. For reproducibility, the source code for
	these experiments is available in the following repository:
	\url{https://github.com/jacoblarock/fake_voices_xai}
    \section{Introduction}
    As the use of generative methods for the creation of synthetic voices becomes more widespread,
    so grows the need for reliable and usable detection methods for the protection of the security
    of people and businesses alike. In particular, the rise of deepfake technology has led to
	concerns about its potential misuse in areas such as politics, entertainment and national
	security by, for instance, potentially allowing malicious actors to create fake audio
	recordings that appear to be genuine statements made by public figures or to create
	genuine-seeming recordings of events that never happened, which could have significant
	consequences if used to spread misinformation, from defaming individuals to creating political
	tension \cite{veerasamy_rising_2022,albahar_deepfakes_2005}.
	\par
	The field of Explainable AI (XAI) shows promise for producing useful and interpretable results
	from such models. Explainable AI refers to the ability of artificial intelligence systems,
	such as machine learning models and neural networks, to provide understandable and
	interpretable explanations for their decisions or predictions \cite{hind_explaining_2019}.
	This means that XAI systems can articulate why they made a particular decision, what factors
	influenced it and how confident they are in their conclusion \cite{hind_explaining_2019}. For
	the detection of audio deepfakes, this could mean that the model can justify its
	classification with feature-based evidence, allowing for verification of the result by the end
	user. Much of the existing research, further discussed in the next section, follows the path
	of producing a model with the best possible benchmarks against datasets of samples, without
	taking into account if the results can be usefully interpreted, while existing explorations
	into making audio deepfake detection explainable have produced results that often require a
	lot of background knowledge to meaningfully interpret. Aside from this, explainability in the
	area of audio deepfake detection remains an open challenge \cite{cuccovillo_open_2022}, one
	even promoted by regulatory bodies such as the European Union with the "right to explain"
	\cite{goodman_european_2017}.
	\par
	In this work, we make a distinction of two main categories of input features for a synthetic
	voice detection model: perceptible and imperceptible features. Perceptible features are
	features that can be perceived by the human ear, often vocal qualities such as jitter, shimmer
	or pitch fluctuation that have a wide range of uses even outside of audio classification such
	as diagnosis of disease \cite{chaiwongyen_deepfake-speech_2023}, while imperceptible features
	are typically out of the range of human hearing, and may not directly reflect a vocal quality,
	otherwise referred to as "speaker-independent" features \cite{liu_hidden--wave_2023}. We use a
	new combination of features for our method, with an emphasis being placed on perceptible
	features, as we believe these to be the most likely to be understood by the layperson.
	\par
	We introduce a hybrid deep-learning model, with the purpose of combining these
	multidimensional features into a single output. The hybrid model consists firstly of a
	component sub-model for each of the input features, in order to address the problem of varying
	dimensionality of the extracted features without the drawbacks of costly transformations in
	the pre-processing phase. The final part of our hybrid model is what we will refer to as the
	terminus model, a model that uses the concatenated outputs of the individual sub-models as
	inputs and distills down to a singular output with a classic multi-layer perceptron
	architecture, allowing for an effective combination of the features into a final singular
	classification result.
	\par
	In order to introduce explainability into our method, we generate explanations using a method
	that first generated the intermediate output data from the sub-models and then applies the
	Local Interpretable Model-agnostic Explanations (LIME) \cite{ribeiro_why_2016} on the terminus
	model, which allows us to, for an individual classification, assess the impact of each input
	feature on the final result through the local explanation, as, due the independence of the
	sub-models from one another, the importance of their outputs directly correlates with the
	importance of their inputs when it comes to the end-classification made by the terminus model.
	\par
	We then stood before the problem of assessing our explanations, more specifically, assessing
	the impact of the input features on the average local explanation, in order to make a
	conclusion about the usefulness or promise of our chosen features for providing understandable
	explanations of our model's classifications. In order to tackle the problem with our local
	explanations, and due to a lack of metrics to assess the aggregated average value of local
	explanations, we also introduce two metrics for such an aggregate evaluation of large number
	of generated LIME explanations, importance and trust, on a per-feature basis, which we use to
	make a conclusion about the value of the selected feature set for the purpose of sensible
	explanations.
	\par
	The contributions of this work are in summary: Our hybrid model-architecture and modified
	LIME-based method for generating explanations of classifications of audio voice samples as
	synthetic or bona-fide and the two metrics that we introduce in order to assess the impact of
	individual features over numerous local explanations.
	\par
	This paper is divided into the following sections:
	\begin{itemize}
		\item §\ref{sec:related} Related Work, where we discuss the most relevant related
			research in the field, in order to create a picture of the current state-of-the-art
		\item §\ref{sec:method} Method, where we describe the general model implementation, as
			well as the implementation of our LIME-based method for generating explanations in
			more detail.
		\item §\ref{sec:experimental_results} Experimental Results, where we discuss how we performed our
			experiments, how our hybrid model performs, as well as use our proposed metrics in
			order to assess the feature impact in the average explanation.
		\item §\ref{sec:discussion} Discussion, where we discuss the implication of the results of
			our experiments, including the potential of the explanations as well as the
			shortcomings and limitations of our method.

@@

	\section{Related Work}
	\label{sec:related}
	Several researchers have already followed similar or relevant directions to that of this work.
	This section will summarize some efforts in previous research divided into categories of
	perceptible and imperceptible features, as well as previous efforts at explainability in this
	field.
	\subsection{Synthetic Voice Detection With Imperceptible Features}
	The most common approach in this field makes use of either learned or hand-crafted
	imperceptible features. Examples include spectrographic features such as the mel-spectrogram
	and its hand-crafted derivative the mel-frequency cepstral coefficients (MFCCs), both of which
	are present in our method due to their widespread successful use.
	\par
	The work of Anagha et al. \cite{anagha_audio_2023} is an example that makes use of
	mel-spectrograms in combination with an architecture based on convolutional neural networks,
	achieving well-performing results on the ASVSpoof2019 dataset \cite{wang_asvspoof_2020}.
	\par
	\sloppy

>>>>

\documentclass{article}
\usepackage{graphicx} % Required for inserting images
\usepackage{biblatex} % Imports biblatex package
\addbibresource{references.bib}
\usepackage{changepage}
\usepackage{parskip}
\usepackage{subfig}
\usepackage{multirow}

@@

    {\begin{table}[h]
			\centering
			\begin{tabular}{lll}
				{Erstprüfer:} & \erstpruefer & \erstprueferMail \\
			\end{tabular}
		\end{table}}     
		\vfill
	\end{titlepage}
    \newpage
\section{Abstract}
\sloppy
With the unprecedented advancement of Generative Artificial Intelligence (GenAI), the threat of voice scams using synthetic voices has become a serious concern across various sectors. Recent efforts have focused on identifying fake voices through handcrafted features, deep learning models, and hybrid approaches. However, most existing methods lack explainability, rendering their predictions non-transparent to users. This paper proposes a novel, interpretable, and transparent method for fake voice identification by introducing a hybrid deep learning model that leverages multiple extracted features. The hybrid model consists of two main components: the first component addresses heterogeneous feature spaces by employing deep convolutional sub-models tailored to individual features, while the second component, the terminus model, utilizes the concatenated representations from the final layers of each sub-model as input. The terminus model follows a typical multi-layer perceptron architecture, enabling effective integration and classification of the diverse feature representations. To enhance interpretability, we decompose the model’s decisions using Local Interpretable Model-agnostic Explanations (LIME), taking advantage of the identical feature representation before the concatenation layers to address challenges related to multidimensional feature representations. To evaluate the features and assess the quality of the generated explanations, we propose two metrics: importance and trust. Extensive experiments are conducted on the In-the-Wild dataset, which is designed to test the generalization capability of synthetic audio detection methods. The experimental results demonstrate that our approach achieves performance comparable to benchmark methods. Furthermore, the results based on our proposed metrics conclude that certain perceptible features demonstrate promise for generating explanations that are meaningful to general users. For reproducibility, the source code for these experiments is available in the following repository: \url{https://github.com/jacoblarock/fake_voices_xai}

\section{Introduction}
Since the use of generative methods for creating synthetic voices has become more widespread, the need for reliable and usable detection methods to protect the security of individuals and businesses has grown. In particular, the rise of deepfake technology has raised concerns about its potential misuse in areas such as politics, entertainment, and national security. For instance, malicious actors could exploit this technology to create fake audio recordings that appear to be genuine statements made by public figures or to fabricate recordings of events that never occurred. Such misuse could have significant consequences, including the spread of misinformation, defamation of individuals, and the escalation of political tensions \cite{veerasamy_rising_2022,albahar_deepfakes_2005}.

There has been considerable attention on identifying audio deepfakes using classical and sophisticated deep learning models~\cite{wang2020deepsonar,becker2024audiomnist,muller_does_2022,yang_robust_2024,wang_asvspoof_2020}. Methods for fake voice identification can be broadly categorized into two classes: methods with handcrafted feature extraction and methods with end-to-end deepfake detectors~\cite{dixit2023review,li2025survey}. In the former, detection approaches first extract various features from the voices, including perceptible features and Mel-spectrograms. These high-dimensional feature values are then passed through complex deep learning models to determine whether the voice is real or synthesized (i.e., fake). End-to-end fake voice identification methods generally optimize the feature extraction process and classification task jointly. However, in recent times, these methods have become more popular than the former category.

The field of Explainable AI (XAI) shows promise for producing useful and interpretable results from such models. Explainable AI refers to the ability of artificial intelligence systems, such as machine learning models and neural networks, to provide understandable and interpretable explanations for their decisions or predictions~\cite{hind_explaining_2019}. This means that XAI systems can articulate why they made a particular decision, what factors influenced it, and how confident they are in their conclusion \cite{hind_explaining_2019}. Although the application of XAI has increased exponentially in other domains, including health informatics, natural language processing, and image processing, its use in fake voice identification is less frequent. One probable reason is that voice consists of signals, and explanations leveraging signals might not be as meaningful compared to other domains. For example, explanations highlighting objects in object identification or relevant words in text sentiment classification can make sense to users.

For the detection of audio deepfakes, this could mean that the model can justify its classification with feature-based evidence, allowing for verification of the result by the end user. Much of the existing research, further discussed in the next section, focuses on producing models with the best possible benchmarks against datasets of samples, without considering whether the results can be usefully interpreted. Existing explorations into making audio deepfake detection explainable have produced results that often require significant background knowledge to interpret meaningfully. Aside from this, explainability in the area of audio deepfake detection remains an open challenge \cite{cuccovillo_open_2022}, one even promoted by regulatory bodies such as the European Union with the "right to explain"~\cite{goodman_european_2017}.

One popular method to detect synthetic voices is to use images consisting of Mel-spectrograms derived from voice signals. Deep learning classifiers then utilize these generated images to predict whether the voice is original or fake. However, applying XAI to such methods is not meaningful because explanations considering pixels as elements would not be understandable to general users. Such explanations can help AI practitioners understand the problem and modify models to achieve higher accuracy. An alternative solution for explaining predictions in fake voice identification is a feature extraction-based approach. In this case, predictions can be explained using understandable features.

In this work, we introduce a hybrid deep learning model leveraging perceptible features extracted from voice signals. The dimensions of each perceptible feature are not the same. Therefore, our hybrid model consists of a component sub-model for each input feature to address the problem of varying dimensionality of the extracted features without the drawbacks of costly transformations in the pre-processing phase. The concatenation of the sub-models' outputs is followed by a terminus model. The terminus model, which uses the concatenated outputs of the individual sub-models as inputs, distills them into a singular output using a classic multi-layer perceptron architecture, allowing for an effective combination of the features into a final classification result.

Since Mel-spectrogram-based representations are not suitable for explaining predictions to users, we make use of perceptible features. Perceptible features are features that can be perceived by the human ear, such as jitter, shimmer, or pitch fluctuation, which have a wide range of uses even outside of audio classification, such as in disease diagnosis \cite{chaiwongyen_deepfake-speech_2023}. To introduce explainability into our method, we generate explanations using the output of each sub-model before the concatenation layers and then apply Local Interpretable Model-agnostic Explanations (LIME) \cite{ribeiro_why_2016} on the terminus model. This allows us to assess the impact of each input feature on the final result through local explanations. Due to the independence of the sub-models, the importance of their outputs directly correlates with the importance of their inputs for the end classification made by the terminus model.

We then faced the problem of assessing our explanations, specifically evaluating the impact of the input features on the average local explanation to determine the usefulness of our chosen features for providing understandable explanations of our model's classifications. Other methods for assessing the global impact of features exist, such as SHAP (SHapley Additive exPlanations) \cite{lundberg2017unified}. However, because such methods use principles different from those of LIME for assessing the impact of features on a global scale, they do not necessarily represent the presence of the features within the average explanation generated by LIME. Due to this, as well as the fact that existing research lacks methods of aggregate evaluation of many LIME explanations, we therefore introduce two metrics: \textit{trust} and \textit{importance}. These metrics allow for an aggregate evaluation of a large number of generated LIME explanations on a per-feature basis, enabling us to assess the usefulness of the features within a given feature set for producing useful and understandable explanations.

% To address this problem and due to a lack of metrics for assessing the aggregated average value of local explanations, we introduce two metrics: \textit{trust} and \textit{importance}. These metrics allow for an aggregate evaluation of a large number of generated LIME explanations on a per-feature basis, enabling us to assess the value of the selected feature set for producing sensible explanations.

We conducted a wide range of experiments to evaluate the performance of our hybrid audio deepfake detection approach on the \textit{In-The-Wild} dataset. The experimental results demonstrate that our model performs effectively compared to state-of-the-art methods. The evaluation of the generated explanations using our proposed metrics also identifies which features have the highest impact on the model's overall predictions. The contributions of this paper are twofold:
\begin{itemize}
\item We propose an interpretable hybrid deep learning model to identify synthetic voices using perceptible features.
% \item Our method can explain predictions, and based on our introduced evaluation metrics, the explanations can highlight the impact of specific features.
\item We propose two metrics in order to perform aggregate evaluations of a large explanations from our LIME-based method, in order to assess how useful the average explanation is.
\end{itemize}

This paper is divided into the following sections: Section \ref{sec:related} presents the most relevant research conducted in this field. Section \ref{sec:method} describes our proposed hybrid deep learning model and the evaluation metrics introduced to judge the generated explanations. The next section, section \ref{sec:experimental_results} presents the experimental setup, results, and performance comparisons. Section \ref{sec:discussion} discusses the implications of the results and limitations of our method. Finally, the last section concludes the paper with possible future work.

%Since the use of generative methods for the creation of synthetic voices becomes more widespread, the need for reliable and usable detection methods for the protection of the security of people and businesses grows. In particular, the rise of deepfake technology has led to concerns about its potential misuse. There are several areas where the misuse occurs including politics, entertainment and national security by, for instance, potentially allowing malicious actors to create fake audio recordings that appear to be genuine statements made by public figures or to create genuine-seeming recordings of events that never happened, which could have significant consequences if used to spread misinformation, from defaming individuals to creating political tension\cite{veerasamy2022rising,albahar_deepfakes_2005}.

%we proposed a hybrid explainable fake voice detection model leveraging perceptible features extracted from the voice signal. Since the dimension of the different features are not same (i.e., multidimensional), we first pass through the every single feature values through individual sub-models. In other words, the hybrid model has first individual sub-model of every features and the concatenation of the similar representation is going through a terminus model. At the output layer of such sub-model, we have similar representation of the features values with same dimension. Next the terminus model process the concatenation of the output from all sub-models.

%we make a distinction of two main categories of input features for a synthetic voice detection model: perceptible and imperceptible features. Perceptible features are features that can be perceived by the human ear, often vocal qualities such as jitter, shimmer or pitch fluctuation that have a wide range of uses even outside of audio classification such as diagnosis of disease \cite{chaiwongyen_deepfake-speech_2023}, while imperceptible features are typically out of the range of human hearing, and may not directly reflect a vocal quality, otherwise referred to as "speaker-independent" features~\cite{liu_hidden--wave_2023}. We use a new combination of features for our method, with an emphasis being placed on perceptible features, as we believe these to be the most likely to be understood by the layperson.

%The contributions of this work are in summary: Our hybrid model-architecture and modified LIME-based method for generating explanations of classifications of audio voice samples as synthetic or bona-fide and the two metrics that we introduce in order to assess the impact of individual features over numerous local explanations.
	%\begin{itemize}
	%	\item §\ref{sec:related} Related Work, where we discuss the most relevant related research in the field, in order to create a picture of the current state-of-the-art
	%	\item §\ref{sec:method} Method, where we describe the general model implementation, as well as the implementation of our LIME-based method for generating explanations in more detail.
	%	\item §\ref{sec:experimental_results} Experimental Results, where we discuss how we performed our experiments, how our hybrid model performs, as well as use our proposed metrics in order to assess the feature impact in the average explanation.
        %\item §\ref{sec:discussion} Discussion, where we discuss the implication of the results of our experiments, including the potential of the explanations as well as the	shortcomings and limitations of our method.
	%\end{itemize}

\section{Related Work} \label{sec:related}

The detection of fake voices has received significant attention in recent years. This section presents relevant research that utilizes hand-crafted features, both perceptible and imperceptible. Additionally, we highlight efforts within the research community to enhance the explainability of fake voice identification methods.

\subsection{Synthetic Voice Detection with Imperceptible Features}

The most common approach in this domain involves using either learned or hand-crafted imperceptible features. Examples include spectrographic features such as mel-spectrograms and their hand-crafted derivatives, such as mel-frequency cepstral coefficients (MFCCs), both of which are widely used due to their effectiveness.

Anagha et al.~\cite{anagha_audio_2023} utilized mel-spectrograms in combination with a convolutional neural network (CNN)-based architecture, achieving strong performance on the ASVSpoof2019 dataset~\cite{wang_asvspoof_2020}. Similarly, Yan et al.~~\cite{yan_initial_2022} leveraged MFCCs alongside other hand-crafted imperceptible features such as linear frequency cepstral coefficients (LFCCs) to classify audio samples as synthetic or authentic. Their approach also identified the vocoder used to generate the sample with near-perfect accuracy on a handcrafted dataset.

\sloppy
Qais et al.~\cite{qais_deepfake_2022} explored Fourier transforms, particularly the short-term Fourier transform (STFT), in combination with mel-spectrogram-based features. Their CNN-based approach demonstrated high accuracy on the ASVSpoof2017 dataset~\cite{delgado_asvspoof_2018}. Müller et al.~\cite{muller_does_2022} introduced the In-the-Wild dataset to assess the generalizability of deepfake voice detection methods. Their evaluation of various features and architectures revealed that existing methods struggle with generalization.

Yang et al.~\cite{yang_robust_2024} conducted a comparative analysis of multiple features and a feature selection method aimed at improving model efficiency. Their experiments, conducted on ASVSpoof2019\cite{wang_asvspoof_2020}, ASV2021~\cite{liu_asvspoof_2023}, and In-the-Wild~\cite{muller_does_2022} datasets, demonstrated the impact of different features on model performance and emphasized the benefits of feature selection and classification fusion techniques.

Ranjan et al.~\cite{ranjan_statnet_2022} proposed a deep convolutional network designed for both spoof detection and source identification. Their evaluation across ASVSpoof2019\cite{wang_asvspoof_2020}, the FOR-Norm dataset~\cite{reimao_for_2019}, and the In-the-Wild dataset~\cite{muller_does_2022} showed high accuracy in individual dataset evaluations. However, cross-dataset evaluations revealed limitations in generalizability. Yi et al.~\cite{yi_audio_2023} provided a comprehensive comparison of features and model architectures for synthetic voice detection. Their study, conducted on ASVSpoof2019~\cite{wang_asvspoof_2020} and In-the-Wild~\cite{muller_does_2022} datasets, reinforced the observation that existing methods struggle with generalization across datasets.

\subsection{Synthetic Voice Detection with Perceptible Features}

Although less explored, several studies have investigated the use of perceptible features for detecting synthetic voices. Some of these features were incorporated into our experiments.

Barrington et al.~\cite{barrington_single_2023} examined the potential of perceptible features in deepfake audio classification, emphasizing their role in improving explainability. However, while they implemented a classifier, they did not develop an explainer. Their findings indicated a performance drop when using perceptible features compared to imperceptible hand-crafted and deep-learning-based features.

Chaiwongyen et al.~\cite{chaiwongyen_contribution_2022,chaiwongyen_deepfake-speech_2023} explored perceptible feature-based classification. Their initial perceptron model, trained and tested on the ADD2022 Challenge dataset~\cite{yi_add_2024}, showed limited performance in 2022. However, with an expanded feature set in 2023, they achieved improved results. Li et al.~\cite{li_comparative_2022} investigated a hybrid approach combining perceptible and imperceptible (referred to as "physical" in their work) features. Their experiments, conducted using various neural networks on the ASV2022 Challenge dataset~\cite{yi_add_2024}, demonstrated that integrating both feature types yielded the best performance, outperforming models trained on only one type of feature.

Efforts have also been made to integrate explainability into synthetic voice detection models. Ge et al.~\cite{ge_explaining_2024} applied the SHAP (SHapley Additive exPlanations) method to analyze feature influence in deepfake audio detection. Using log-scaled power spectrograms as input features, they trained and tested their model on the ASV2019 dataset~\cite{wang_asvspoof_2020}. Their approach enabled graphical representation of feature importance on spectrograms and a global summary of SHAP values.

Haq et al.~\cite{haq_multimodal_2023} proposed an explainable approach by leveraging emotional state changes as input features. Their method visually represented "unlikely" emotional shifts to enhance interpretability for end users. By combining fake video and fake audio classifiers, they produced a final classification for video samples with audio. Their model, tested on the Presidential Deepfake dataset~\cite{sankaranarayanan_presidential_nodate}, achieved superior results compared to existing benchmarks.

%\section{Related Work}	\label{sec:related}
%There has been a significant attention to detect fake voices in recent years. This section presents related research which makes use of hand-crafted features, both perceptible and imperceptible features. In the end, we present the effort of the research community to make fake voice identification methods explainable.

%\subsection{Synthetic Voice Detection with Imperceptible Features}
%The most common approach in this field makes use of either learned or hand-crafted imperceptible features. Examples include spectrographic features such as the mel-spectrogram and its hand-crafted derivative the mel-frequency cepstral coefficients (MFCCs), both of which are present in our method due to their widespread successful use.

%The work of Anagha et al. \cite{anagha_audio_2023} is an example that makes use of mel-spectrograms in combination with an architecture based on convolutional neural networks, achieving well-performing results on the ASVSpoof2019 dataset \cite{wang_asvspoof_2020}. A further work from Yan et al. \cite{yan_initial_2022} makes use of MFCCs, in combination with other hand-crafted imperceptible features such as linear frequency cepstral coefficients (LFCCs) to not only classify audio samples as synthetic or authentic, but also to correctly classify the vocoder used to produce the sample with nearly perfect accuracy on a handcrafted dataset.

%Other researchers, such as in the work of Qais et al. \cite{qais_deepfake_2022} make use of Fourier transforms, such as the short term Fourier transform (STFT). The aforementioned example uses these features combined with others such as the aforementioned mel-spectrogram based features. They also pair these features with a convolutional architecture and achieve notably accurate results on the ASVSpoof2017 dataset \cite{delgado_asvspoof_2018}. With their work, Müller et al. \cite{muller_does_2022} introduce the In-the-Wild dataset, in an attempt to answer the question if -- and to what extent -- previously proposed methods of deepfake voice detection can generalize. They test a range of features and architectures on their dataset, concluding that there is still work to be done in the field, due to a generally lackluster performance.

%Yang et al. \cite{yang_robust_2024} demonstrate a comparison of multiple features, as well as a feature selection method, with the goal of maximizing model efficiency. They run experiments with several hand-crafted features as well as learned features. They train and test using three datasets, ASVSpoof2019 \cite{wang_asvspoof_2020}, ASV2021 \cite{liu_asvspoof_2023} and In-the-Wild \cite{muller_does_2022}, presenting their results for each feature, as well as the results of their selection and classification fusion methods.

%The work of Ranjan et al. \cite{ranjan_statnet_2022} introduces a deep convolutional network that is designed not only for spoof detection, but also for the purpose of source detection. They perform experiments on several datasets, including ASVSpoof2019 \cite{wang_asvspoof_2020}, the FOR-Norm dataset \cite{reimao_for_2019} and the In-the-Wild dataset \cite{muller_does_2022}. In the experiments run on each dataset, their method was able to achieve high evaluation performance, but they were not able to achieve high-performance in cross-dataset evaluations, indicating a lack of generalizability.

%Yi et al. \cite{yi_audio_2023} present in their work a comprehensive comparison of features and model architectures used for detection methods, giving results from both the ASVSpoof2019 \cite{wang_asvspoof_2020} and In-the-Wild \cite{muller_does_2022} datasets, once again demonstrating a lack of generalizability of existing methods, consistent across various kinds of features and architectures.

%\subsection{Synthetic Voice Detection With Perceptible Features}
%There is also a smaller but still notable body of work encompassing the use of perceptible features for the detection of synthetic voices, making use of a variety of features, some of which were selected for use in our experiments.

%The work of Barrington et al.~\cite{barrington_single_2023} explores the idea of classification using perceptible features, highlighting their potential for improving explainability in the space of deepfake audio detection, although they only implement a classifier and not an explainer. They do, however, also demonstrate a drop in performance when using perceptible features in comparison with imperceptible hand-crafted and deep-learning features in their experiments, which were performed on a combination of synthetic and real audio datasets.

%Chaiwongyen et al.~\cite{chaiwongyen_contribution_2022,chaiwongyen_deepfake-speech_2023} also approach using perceptible features for classification in their works. They developed a perceptron architecture with a single hidden layer and trained and tested using the dataset of the ADD2022 Challenge~\cite{yi_add_2024}, resulting in lackluster performance in 2022, with better results using an expanded and improved feature set in 2023. Li et al.~\cite{li_comparative_2022} also approach perceptible features in their work, combining them with imperceptible (referred to as physical in the work) features and using various neural networks in their experiments. With their combination of perceptible and imperceptible features, they were able to produce the best performance results of their experiments in comparison to experiments run with only perceptible or imperceptible features, which were performed as well on the dataset of the ASV2022 Challenge~\cite{yi_add_2024}.

%\subsection{Explainable Models}
%There have also been some efforts in the previous research to pair synthetic voice detection models with explainable methods.

%One previous example of an implementation of an explainable model for use in audio deepfake detection came from Ge et al.~\cite{ge_explaining_2024}, who explain feature influence on models using the SHAP (SHapley Additive exPlanations) method. They apply this method using log-scaled power spectrograms as a feature, training and testing using the datasets from the ASV2019 challenge~\cite{wang_asvspoof_2020}, they are able to determine and graphically represent areas of importance on the spectrogram, as well as globally summarize the SHAP-values.

%Another example is presented by Haq et al.~\cite{haq_multimodal_2023}, who use the changes in emotional state as an input feature and represent "unlikely" changes on a graph so that it can be understood by an eventual end user. In this case, they achieved their results by combining the output of fake video and fake audio classifiers in order to produce a final classification for a video sample with audio. Testing against the presidential deepfake dataset \cite{sankaranarayanan_presidential_nodate}, they achieve impressive results in comparison to the standing benchmark on the dataset at the time.
 \begin{figure}[t]
    \centering
    \includegraphics[width=0.75\linewidth]{images/Picture1.png}
    \caption{A high level building-block of our proposed hybrid fake voice identification model, where \(m\) and \(k\) represent the dimensionality of the given input feature.}
    \label{fig:overview}
\end{figure}
\section{Method}\label{sec:method}
Our proposed fake voice identification method consists of three major components including feature extractor, hybrid detection model, and the generation of explanations by modified LIME. After extracting features, we pass them individually through sub-models to overcome the multidimensionality problem. Then the concatenation of the outputs from the output layers of the sub-models are then passed through the \textit{terminus} models. For generating explanations, we make use of the feature values returned from the output layers of the sub-models, which has same representation of every feature.

The high-level buidling-blocks of our hybrid interpretable deep fake voice identification model are presented in Fig.~\ref{fig:overview}.

\subsection{Feature Extraction}
The extracted features from audio samples can be categorise in two different class: perceptible and imperceptible features. The perceptible features are features that can be perceived by the human ear, often vocal qualities such as jitter, shimmer or pitch fluctuation that have a wide range of uses even outside of audio classification such as diagnosis of disease~\cite{chaiwongyen_deepfake-speech_2023}, while imperceptible features are typically out of the range of human hearing, and may not directly reflect a vocal quality, otherwise referred to as ``speaker-independent'' features~\cite{liu_hidden--wave_2023}. In order to increase the likelihood that the explanations are useful and understandable to the end user, we focused on using multiple perceptible features as input to the classifier. However, we used two imperceptible features with the hypothesis, based on previous research~\cite{barrington_single_2023,chaiwongyen_contribution_2022,chaiwongyen_deepfake-speech_2023,li_comparative_2022}, that they would positively increase model performance. We extracted the following perceptible features:

%Several features were used for the final version of the black-box model. Some are features widely used in research involving the detection of synthetic while others are more specific and are cited accordingly. The features that are not summarized for a whole sample are extracted using a sliding window method, preventing a loss of fidelity that can be caused by compression of the feature to a standard size. The features we considered perceptible are as follows:

\begin{itemize}
\item \textbf{Harmonic-to-noise ratios (HNRs)}: Inspired by previous research~\cite{chaiwongyen_contribution_2022,chaiwongyen_deepfake-speech_2023,		li_comparative_2022}, we extracted this features in a sliding window fashion instead of on the whole file. The HNRs are the ratios of the strengths of harmonic frequencies to the strengths of "noise" and the total strength outside the harmonic frequencies. Unlike previous research~\cite{chaiwongyen_contribution_2022, chaiwongyen_deepfake-speech_2023}, we computed this ratio for each fundamental frequency length, instead of using one value for the whole sample. Given that \(\gamma_{i}\) is the harmonic	energy in a given fundamental frequency cycle and \(\iota_{i}\) is the residual energy in a given fundamental frequency cycle, the HNR at that cycle \(hnr_{i}\) is calculated as follows:

\begin{equation}
    hnr_{i} = 20log\frac{\gamma_{i}}{\iota_{i}}.
\end{equation}

\item \textbf{Fundamental frequency lengths}: The fundamental frequency lengths, f0 length, are the lengths of every fundamental frequency cycle in the sample. This has also been previously employed and found effective in synthetic audio detection~\cite{xue_audio_2022}. The output is one-dimensional with time as the axis.

\item \textbf{Onset strength}: This perceptible feature represents the strengths of each onset in the audio sample, where an onset is point where there is a sudden rise in energy across the audio spectrum~\cite{li_comparative_2022}. This results in a one-dimensional output with time as the axis.

\item \textbf{Intensity}: Intensity is the total power at each point in the audio sample, given in \textit{db}. This is calculated by creating a Fourrier transformation of the sample and then summing across the frequency-axis for every point on the time-axis. Resulting in a one-dimensional output~\cite{li_comparative_2022}.

\item \textbf{Pitch-fluctuations}: The pitch fluctuations are calculated for every sample in the audio as the difference between pitch at the given point and the pitch at the previous point. Pitch is estimated based on the maximum power harmonics. This feature is similar to the use of summarized pitch fluctuations in previous research~\cite{khanjani_learning_2023}. This feature is also one-dimensional with time as the axis. Given that \(H_{i}\) is the set of harmonic frequencies at fundamental frequency cycle \(i\) with \(h_{i} \in H_{i}\) as a frequency of the set and \(s(h_{i})\) is the power of a given harmonic frequency at cycle \(i\), the pitch can be estimated as follows:

\begin{equation}
    p_{i} = s(max(H_{i})).
\end{equation}

Given an offset \(x\), the pitch fluctuation at cycle \(i\), \(pf_{i}\), can be calculated as follows:

\begin{equation}
    pf_{i} = p_{i}-p_{i-x}.
\end{equation}

\item \textbf{Jitter features}: Jitter~\cite{chaiwongyen_deepfake-speech_2023} measures the absolute variations in fundamental frequency cycles in comparison to the nearest \(x\) Neighbors. The jitter of a sample in relation to the nearest \(x\) samples can be described as follows:
\begin{equation}
    jitter(x) = \frac{ \frac{1}{N-1}\sum_{i=1}^{N-1}|T_{i}
						(\frac{1}{x}\sum_{n=i-m}^{i+m}T_{n})|}
					{\frac{1}{N}\sum_{i=1}^{N}T_{i}}, 
\end{equation}

where \(T_{i}\) represents the extracted fundamental frequency length at cycle \(i\), \(N\) is the number of fundamental frequency periods and \(m\) is \(\lfloor \frac{x}{2} \rfloor\).

\item \textbf{Shimmer features}: Similar to Jitter, shimmer instead make use of the amplitudes at each fundamental frequency period. The purpose is to capture irregular vocal fold vibrations which may be an indication but not a guarantee of a synthetic voice. The shimmer of a sample in comparison to the nearest \(x\) fundamental frequency cycles is calculated as follows~\cite{chaiwongyen_deepfake-speech_2023}:
\begin{equation}
    shimmer(x) = \frac{ \frac{1}{N-1}\sum_{i=1}^{N-1}|A_{i}
						(\frac{1}{x}\sum_{n=i-m}^{i+m}A_{n})|}
					{\frac{1}{N}\sum_{i=1}^{N}A_{i}},
\end{equation}
where \(A_{i}\) is the amplitude at cycle \(i\), \(N\) is the number of fundamental frequency cycles in the sample and \(m\) is once again \(\lfloor \frac{x}{2} \rfloor\)

@@

We also chose to include two imperceptible features in our hybrid model, due to the performance-enhancing impact combining with the perceptible features~\cite{chaiwongyen_deepfake-speech_2023}. These features are the mel-spectrogram and their derivative mel-frequency cepstral coefficients. We select these two features as they are well established in the research and have demonstrated consistent high performance in existing state-of-the-art research~\cite{qais_deepfake_2022,anagha_audio_2023,fathan_mel-spectrogram_2022,altalahin_unmasking_2023,hamza_deepfake_2022,yan_initial_2022}. They allow for a	spectrographic representation and analysis of an audio sample transformed in a way to better represent human perception of frequencies~\cite{qais_deepfake_2022}.

\subsection{Hybrid Fake Voice Detection Model}
%We created, trained and evaluated the model used for our experiments using the tools available in the TensorFlow and Keras libraries \cite{tensorflow2015-whitepaper}. This section will go into further detail about the experimental setup for the black-box model as well as the way in which explanations were created and summarized.
